Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-01-22 manuscript.Rnw
Edits to manuscript, captions. Moved Table 1 to SI1

\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \citep{friend2014}, predictions of global carbon storage \citep{ise2010}, and the CO\textsubscript{2} response of net primary production \citep{dekauwe2014}. Although GVMs typically predict that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves. We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}. Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}. Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes. While GVMs predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.

@@

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = \textrm{LMF} / \textrm{LMA}$ (Fig. 2c and S3).

@@

The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs. To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFT and plant height (and the quadratic term) as predictors (and all interactions). We then replaced PFT with LMA, and found that LMA alone could explain almost as much variation in LMF as PFT ($R^2$ = 0.74 with PFT vs. 0.62 with LMA) and $M_F/A_S$ ($R^2$ = 0.29 vs. 0.28). This is surprising because LMA only captures aspects of leaf morphology. These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.

Many GVMs assume that climate influences biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}). We were thus interested to know whether additional variation in biomass partitioning could be explained by climate, as a basis for model formulation. Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site. Thus, we decomposed LMF as the product,

@@

An advantage of eq. \ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1-2) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Tables 1-2). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

@@

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig. S5). It is thus pertinent to understand whether we can move beyond a PFT-based classification by considering which traits affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked. We found that PFTs differed in LMF, and this was proportional to LMA. It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions. A first approximation within GVMs would be to keep leaf area to stem cross-sectional area constant for a given PFT, and vary biomass allocation accordingly. LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass. An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs.

More fundamentally, our results suggest PFTs defined on the basis of two key traits, LMA and $A_F/A_S$ would better capture patterns of biomass partitioning than existing groups, which only account for differences in one of the two traits identified here. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we used different subsets of the data. Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}. Stem cross-sectional area was measured either at breast height (ca. 1.3m) and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using non-linear regression (see~\ref{sec:predas}).

We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height. In all fitted GAMs, we used species-dataset combination as a random effect. We used the \emph{mgcv} package in R \citep{wood2006}. All variables were log-transformed prior to analysis.

Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated. For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect. We calculated the $R^2$ for linear mixed-effects models as reported by for the fixed effects only \citep{nakagawa2013}.

@@

<<results='asis'>>=
print(xtable(table_varpart2, label="varpart2",
             caption="\\textbf{Explained variance in the ratio of leaf mass to basal stem area, and leaf area to basal stem area.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect. The $r^2$ shown is that explained by the fixed effects only (H = plant height, B = biome, PFT = plant functional type). All fixed effects were highly significant (P \\textless 0.001). In addition to the fixed effects shown, all interactions were added to each of the models."),
      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{\textbf{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning}. a) LMF for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. No leaf area estimates were available for boreal deciduous angiosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and LMF within each PFT. Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figure3}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
    \caption{\textbf{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.} Shown are histograms (as probability density functions) of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types. Arrows indicate means (same as in Fig.~\ref{fig:figure2}).}
    \label{fig:figures5}
\end{figure}

@@

Most leading GVMs predict that biomass partitioning between leaves and woody stems differs between major plant functional types. However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL). We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them (See Methods \ref{sec:partterm} for use of the terms 'partitioning' and 'allocation'). We base this classification on a recent model intercomparison study \cite{dekauwe2014}. We then briefly summarize how current GVMs include effects of climate on biomass partitioning.

@@

In this class of models, some allometric (or other size-dependent) relationship between biomass components is used to calculate biomass allocation in each timestep. For example, in ED2 \cite{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration. Biomass allocation is then calculated at each timestep to satisfy these equations. Other models use leaf area as a constraint. For example, LPJ \cite{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation. Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).

@@

Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \cite{valentinemakela2012,franklin2012}. The SDGVM model \cite{woodward1995} optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy. As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.

>>>>

\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \citep{friend2014}, predictions of global carbon storage \citep{ise2010}, and the CO\textsubscript{2} response of net primary production \citep{dekauwe2014}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves. We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}. Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes. While GVMs assume differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.

@@

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. 2c and S3).

@@

The above results highlight how LMA -- a central trait of the leaf economics spectrum (LES)\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs. To assess whether LMA could replace PFT when predicting biomass partitioning, we fit a linear mixed-effects model to LMF and $M_F/A_S$ using PFT and plant height (and the quadratic term) as predictors (and all interactions). We then replaced PFT with LMA, and found that the model with LMA could explain almost as much variation in LMF as the model with PFT ($R^2$ = 0.74 with PFT vs. 0.62 with LMA) and $M_F/A_S$ ($R^2$ = 0.29 vs. 0.28). This is surprising because LMA only captures aspects of leaf morphology, in indirectly leaf lifespan and photosynthetic capacity via the LES\citep{wrightetal2004}. These results suggest promising avenues for parameterizing and simplifying biomass partitioning routines in GVMs.

Many GVMs assume that climate influences biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}). We were thus interested to know whether additional variation in biomass partitioning could be explained by climate. Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons across climatic gradients, due to the small sample size within each species or site. Thus, we decomposed LMF as the product,

@@

An advantage of eq. \ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1 and S1) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the term $M_F/A_S$ exhibits comparable differences between PFTs to LMF, but unlike LMF, $M_F/A_S$ is nearly independent of plant height (Table 1 and S1). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and sites.

@@

We found that LMF differs predictably among PFTs, however there was considerable variation in all studied variables between species within PFTs (Fig. S5). It is thus pertinent to understand whether we can move beyond a PFT-based classification by considering other traits that affect biomass partitioning. To do this, we decomposed $M_F/A_S$ into LMA ($M_F/A_F$), and the ratio of leaf area to stem cross-sectional area ($A_F/A_S$) (Eq. \ref{eq:M2}).

@@

Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked. We found that PFTs differed in LMF, and this was proportional to LMA. It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions. Based on our findings, a first approximation within GVMs would be to assume leaf area to stem cross-sectional area a parameter that does not vary between PFTs, and vary biomass allocation accordingly. LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem area. Some models already incorporate a similar algorithm (see Methods~\ref{sec:gvmsummary}), while other algorithms may be tuned to yield similar patterns between PFTs as we have presented. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) \citep{Falster2015}, including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_S$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we used different subsets of the data. Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4}. We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature and precipitation from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}).

@@

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area using published conversion factors \citep{barclay2000}, which is most appropriate for comparison to flat leaves \citep{lang1991, chen1992}. Stem cross-sectional area was measured either at breast height (ca. 1.3m) and/or at the base of the plant. For the subset of the data where both were measured, we estimated basal stem area ($A_S$) from breast height stem area ($A_{Sbh}$) using a non-linear regression model (see~\ref{sec:predas}).

We used generalized additive models (GAM) to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as LMF at a common plant height. In all fitted GAMs, we used species-dataset combination as a random effect. We used the \emph{mgcv} package in R \citep{wood2006}. All variables (except MAP and MAT) were log-transformed prior to analysis.

Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated. For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect. We calculated the $R^2$ for linear mixed-effects models for the fixed effects only \citep{nakagawa2013}.

@@

<<results='asis'>>=
print(xtable(table_varpart2, label="varpart2",
             caption="\\textbf{Explained variance by plant functional type (PFT), plant height (H) and biome (B), for five whole-plant variables.} Each variable was added to a linear mixed-effects model, using species within dataset as a random effect. The $R^2$ shown is that explained by the fixed effects only. All fixed effects were highly significant (P \\textless 0.001). In addition to the fixed effects shown, all interactions were added to each of the models."),
      include.rownames=FALSE, caption.placement="top", sanitize.text.function=function(x){x})
@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{\textbf{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning}. a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and $M_F/A_S$ within each PFT. Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figure3}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
    \caption{\textbf{Plant functional types diverge strongly in mass-based partitioning, but converge in area-based partitioning.} Shown are histograms (as probability density functions) of leaf mass per area (an index of leaf construction cost) ($M_F/A_F$), leaf mass per unit basal stem area ($M_F/A_S$) and leaf area per unit basal stem area ($A_F/M_S$) grouped by the three major plant functional types. Arrows indicate means by PFT.}
    \label{fig:figures5}
\end{figure}

@@

Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types. However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL). Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning. We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them. We base this classification on a recent model intercomparison study \cite{dekauwe2014}. Note that we use 'partioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods \ref{sec:partterm}). Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.

@@

In this class of models,  allometric (or other size-dependent) relationships between biomass components are used to calculate biomass allocation in each timestep. For example, in ED2 \cite{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters, estimated through model calibration. Biomass allocation is then calculated at each timestep to satisfy these equations. Other models use leaf area as a constraint. For example, LPJ \cite{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation. Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).

@@

Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \cite{valentinemakela2012,franklin2012}. One example of a global model is SDGVM \cite{woodward1995}, which optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy. As with LPJ described above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.
